---
title: "Data Acquisition"
author: "Rick Forest"
date: today

execute:
  echo: true
  warning: false
  message: false

format:
  html:
    toc: true
    number-sections: false
    code-fold: true
    code-summary: "Show code"
    code-tools: true
---

**Objective:** Acquire the ERA5 temperature, precipitation, and land-sea-mask data required for the Köppen–Trewartha climate-classification change analysis. This notebook downloads the source data for the REF, CSN, and RCI climate intervals, records acquisition metadata, and creates a catalog of the files used in subsequent analysis.

## Initialization

```{r}
#| label: Libraries

suppressPackageStartupMessages({
  library(tidyverse)
  library(gt)
  library(fs)
  library(ecmwfr)
})

```

```{r}
#| label: Directories

dir_raw <- path("data", "raw")
dir_transformed <- path("data", "transformed")
dir_cache <- path("data", "cache")
dir_reference <- path("data", "reference")
dir_derived <- path("data", "derived")

dir_create(c(
  dir_raw,
  dir_transformed,
  dir_cache,
  dir_reference,
  dir_derived
))

log_file_path <- path(dir_raw, "acquisition_log.csv")

```

```{r}
#| label: Control

# The three climate intervals can be acquired independently. During a routine
# update, normally only acquire_recent_interval needs to be TRUE after updating
# the RCI end year in R/climate-periods.R.
acquire_reference_period <- FALSE
acquire_standard_normal <- FALSE
acquire_recent_interval <- FALSE

# The ERA5 land-sea mask is static for this workflow and normally needs to be
# downloaded only once.
acquire_land_sea_mask <- FALSE

# Existing files are retained by default so reruns do not repeat large downloads.
overwrite_existing <- FALSE

```

## Project Software Environment

```{r}
#| label: software-versions

quarto_version <- system2("quarto", "--version", stdout = TRUE)[1]
positron_version <- "2026.8.1" # Update manually when Positron is upgraded

software_versions <- data.frame(
  category = c(
    "Platform", "Platform", "Platform",
    "Package", "Package", "Package",
    "Package", "Package", "Package",
    "Package", "Package"
  ),
  software = c(
    "R", "Quarto", "Positron",
    "tidyverse", "gt", "fs",
    "ecmwfr", "terra", "arrow",
    "sf","patchwork"
  ),
  version = c(
    as.character(getRversion()),
    quarto_version,
    positron_version,
    as.character(packageVersion("tidyverse")),
    as.character(packageVersion("gt")),
    as.character(packageVersion("fs")),
    as.character(packageVersion("ecmwfr")),
    as.character(packageVersion("terra")),
    as.character(packageVersion("arrow")),
    as.character(packageVersion("sf")),
    as.character(packageVersion("patchwork"))
  ),
  used_in = c(
    "01–03","01–03","Development",
    "01–03","01–03","01–03",
    "01","01","01–03",
    "02","03"
  ),
  stringsAsFactors = FALSE
)

```

```{r}
#| label: software-versions-table
#| echo: false

software_versions |>
  gt(
    groupname_col = "category"
  ) |>
  cols_label(
    software = "Software",
    version = "Version",
    used_in = "Used In"
  ) |>
  cols_align(
    align = "center",
    columns = c(version, used_in)
  ) |>
  cols_width(
    software ~ px(180),
    version ~ px(120),
    used_in ~ px(110)
  ) |>
  tab_header(
    title = "Project Software Environment",
    subtitle = "Platform software and R packages used in the workflow"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  tab_source_note(
    source_note = md(
      "**Used In:** 01 = Data Acquisition, 02 = Data Aggregation,
                    03 = Determinant Calculation, 04 = Climate Classification,
                    05 = Transition Analysis, 06 = Transition Mapping"
    )
  )

```

## Climate Periods

The analysis uses three non-overlapping climate intervals to provide a consistent framework for comparing long-term and recent climate conditions and their resulting classifications.

```{r}
#| label: climate-periods

source("R/climate-periods.R")

climate_periods |>
  transmute(
    `Period` = period_name,
    `Short Label` = period_label,
    `Abbreviation` = period_abbr,
    `Years` = sprintf("%d–%d", start_year, end_year)
  ) |>
  gt()

reference_period <- climate_periods |>
  filter(period_id == "reference_period")

standard_normal <- climate_periods |>
  filter(period_id == "standard_normal")

recent_interval <- climate_periods |>
  filter(period_id == "recent_interval")
```

## Classification Reference

```{r}
#| label: classification-group-reference

ktc_group <- c(
  "A", "B", "C", "D", "E", "F"
)

ktc_group_labels <- c(
  A = "Tropical",
  B = "Dry",
  C = "Subtropical",
  D = "Temperate",
  E = "Boreal",
  F = "Polar"
)

ktc_group_hex <- c(
  A = "#cf5a55",
  B = "#e79a6a",
  C = "#1f9d6d",
  D = "#41b7c8",
  E = "#2a6bb4",
  F = "#a285c9"
)

ktc_group_palette <- tibble(
  ktc = ktc_group,
  hex = ktc_group_hex,
  lbl = ktc_group_labels
)

write_csv(
  ktc_group_palette,
  path(
    dir_reference,
    "ktc_group_codes.csv"
  )
)

```

```{r}
#| label: classification-type-reference

ktc_type <- c(
  "Ar", "Aw",
  "BW", "BS",
  "Cs", "Cw", "Cf",
  "Do", "Dc",
  "Eo", "Ec",
  "Ft", "Fi"
)

ktc_type_labels <- c(
  Ar = "Tropical Rainforest",
  Aw = "Tropical Savanna (dry season)",
  BW = "Desert (Arid)",
  BS = "Steppe (Semiarid)",
  Cs = "Mediterranean (dry summer)",
  Cw = "Subtropical (dry winter)",
  Cf = "Subtropical (no dry season)",
  Do = "Temperate (Oceanic)",
  Dc = "Temperate (Continental)",
  Eo = "Boreal (Oceanic)",
  Ec = "Boreal (Continental)",
  Ft = "Tundra",
  Fi = "Ice Cap"
)

ktc_type_hex <- c(
  Ar = "#85080c",
  Aw = "#cb1e0a",
  BW = "#fede49",
  BS = "#f09036",
  Cs = "#38d98f",
  Cw = "#1f9d6d",
  Cf = "#13724f",
  Do = "#02add7",
  Dc = "#0f74bb",
  Eo = "#1e66c1",
  Ec = "#103e7b",
  Ft = "#9c84c7",
  Fi = "#60468d"
)

ktc_type_palette <- tibble(
  ktc = ktc_type,
  hex = ktc_type_hex,
  lbl = ktc_type_labels
)

write_csv(
  ktc_type_palette,
  path(
    dir_reference,
    "ktc_type_codes.csv"
  )
)

```

## Functions

```{r}
#| label: acquisition-functions

fn_download_request <- function(request, output_dir, overwrite = FALSE) {
  output_path <- path(output_dir, request$target)

  if (file_exists(output_path) && !overwrite) {
    return(tibble(
      file_name = path_file(output_path),
      variable_name = request$variable,
      status = "existing",
      file_size_mb = round(as.numeric(file_info(output_path)$size) / 1024^2, 2),
      elapsed_mmss = "00:00"
    ))
  }

  if (file_exists(output_path) && overwrite) {
    file_delete(output_path)
  }

  start_time <- Sys.time()

  downloaded_path <- ecmwfr::wf_request(
    request = request,
    transfer = TRUE,
    path = output_dir,
    verbose = TRUE
  )

  elapsed_seconds <- as.numeric(
    difftime(Sys.time(), start_time, units = "secs")
  )

  tibble(
    file_name = path_file(downloaded_path),
    variable_name = request$variable,
    status = "downloaded",
    file_size_mb = round(as.numeric(file_info(downloaded_path)$size) / 1024^2, 2),
    elapsed_mmss = sprintf(
      "%02d:%02d",
      floor(elapsed_seconds / 60),
      round(elapsed_seconds %% 60)
    )
  )
}

fn_climate_request <- function(variable, beg, end, target) {
  list(
    dataset_short_name = "reanalysis-era5-single-levels-monthly-means",
    product_type = "monthly_averaged_reanalysis",
    variable = variable,
    year = sprintf("%04d", beg:end),
    month = sprintf("%02d", 1:12),
    time = "00:00",
    data_format = "netcdf",
    download_format = "unarchived",
    target = target
  )
}

fn_mask_request <- function(target) {
  # Request the native global ERA5 grid. This keeps the mask aligned with the
  # climate files while the large monthly rasters remain on 0..360 longitude.
  list(
    dataset_short_name = "reanalysis-era5-single-levels",
    product_type = "reanalysis",
    variable = "land_sea_mask",
    year = "2000",
    month = "01",
    day = "01",
    time = "00:00",
    data_format = "netcdf",
    download_format = "unarchived",
    target = target
  )
}

fn_period_requests <- function(beg, end) {
  list(
    fn_climate_request(
      "2m_temperature",
      beg,
      end,
      sprintf("t_avg_%d-%d_ERA5.nc", beg, end)
    ),
    fn_climate_request(
      "total_precipitation",
      beg,
      end,
      sprintf("p_avg_%d-%d_ERA5.nc", beg, end)
    )
  )
}
```

## Acquire Data

```{r}
#| label: acquire-data

requests <- list()

if (acquire_reference_period) {
  requests <- c(
    requests,
    fn_period_requests(reference_period$start_year, reference_period$end_year)
  )
}

if (acquire_standard_normal) {
  requests <- c(
    requests,
    fn_period_requests(standard_normal$start_year, standard_normal$end_year)
  )
}

if (acquire_recent_interval) {
  requests <- c(
    requests,
    fn_period_requests(recent_interval$start_year, recent_interval$end_year)
  )
}

if (acquire_land_sea_mask) {
  requests <- c(
    requests,
    list(fn_mask_request("land_sea_mask_ERA5_025deg.nc"))
  )
}

if (length(requests) > 0) {
  log_df <- bind_rows(lapply(
    requests,
    fn_download_request,
    output_dir = dir_raw ,
    overwrite = overwrite_existing
  ))

  log_df <- log_df |>
    mutate(acquisition_datetime = Sys.time())

  readr::write_csv(log_df, log_file_path)
} else {
  message("No acquisitions requested. Set one or more acquisition controls to TRUE.")
}
```

```{r}
#| label: acquire-coastline

coastline_path <- path(
  dir_reference,
  "global-coastline.gpkg"
)

if (!file_exists(coastline_path)) {
  coastline_sf <- rnaturalearth::ne_download(
    scale = 10,
    type = "coastline",
    category = "physical",
    returnclass = "sf"
  ) |>
    st_make_valid()

  st_write(
    coastline_sf,
    coastline_path,
    delete_dsn = TRUE,
    quiet = TRUE
  )
}
```

## Acquisition Summary

```{r}
#| label: update-catalog

climate_variables <- tibble::tribble(
  ~climate_variable, ~variable_name,
  "t",               "Temperature",
  "p",               "Precipitation"
)

climate_file_catalog <- tidyr::crossing(
  climate_periods,
  climate_variables
) |>
  mutate(
    dataset = paste(
      variable_name,
      period_label
    ),
    source = "ERA5",
    format = "NetCDF",
    filename = sprintf(
      "%s_avg_%d-%d_ERA5.nc",
      climate_variable,
      start_year,
      end_year
    ),
    local_path = path(
      dir_raw,
      filename
    )
  )

spatial_file_catalog <- tibble(
  period_id = NA_character_,
  period_name = NA_character_,
  period_label = NA_character_,
  period_abbr = NA_character_,
  start_year = NA_integer_,
  end_year = NA_integer_,
  climate_variable = "lsm",
  variable_name = "Land-sea mask",
  dataset = "ERA5 Land-Sea Mask",
  source = "ERA5",
  format = "NetCDF",
  filename = "land_sea_mask_ERA5_025deg.nc",
  local_path = path(
    dir_raw,
    filename
  )
)

file_catalog <- bind_rows(
  climate_file_catalog,
  spatial_file_catalog
) |>
  filter(
    file_exists(local_path)
  )

file_metadata <- file_info(
  file_catalog$local_path
)

file_catalog <- file_catalog |>
  mutate(
    directory = path_dir(local_path),
    acquisition_datetime = format(
      file_metadata$modification_time,
      "%Y-%m-%d %H:%M"
    ),
    file_size_mib = round(
      as.numeric(file_metadata$size) / 1024^2,
      1
    )
  ) |>
  arrange(
    start_year,
    climate_variable
  )

```


```{r}
#| label: acquisition-summary
#| echo: false

file_catalog |>
  mutate(
    climate_period = case_when(
      is.na(period_abbr) ~ "Static",
      TRUE ~ period_abbr
    )
  ) |>
  select(
    dataset,
    climate_period,
    filename,
    format,
    acquisition_datetime,
    file_size_mib
  ) |>
  gt() |>
  cols_label(
    dataset = "Dataset",
    climate_period = "Climate Period",
    filename = "Filename",
    format = "Format",
    acquisition_datetime = "Acquired",
    file_size_mib = "Size (MiB)"
  ) |>
  fmt_number(
    columns = file_size_mib,
    decimals = 1
  ) |>
  cols_align(
    align = "center",
    columns = c(
      climate_period,
      format,
      acquisition_datetime,
      file_size_mib
    )
  ) |>
  cols_width(
    dataset ~ px(220),
    climate_period ~ px(110),
    filename ~ px(250),
    format ~ px(90),
    acquisition_datetime ~ px(135),
    file_size_mib ~ px(95)
  ) |>
  tab_header(
    title = "Project Data Catalog",
    subtitle = "ERA5 source data acquired for the analysis"
  )
```