---
title: "Determinant Calculation"
author: "Rick Forest"
date: today

execute:
  echo: true
  warning: false
  message: false

format:
  html:
    toc: true
    number-sections: false
    code-fold: true
    code-summary: "Show code"
    code-tools: true
---

**Objective:** Calculate the temperature, precipitation, and aridity determinants used by the Köppen–Trewartha climate-classification analysis for the REF, CSN, and RCI climate intervals. This notebook combines the transformed ERA5 climatologies, derives the classification determinants for each grid cell and interval, saves a unified determinant dataset for subsequent climate classification, and summarizes the area-weighted results.

## Initialization

```{r}
#| label: Libraries

suppressPackageStartupMessages({
  library(tidyverse)
  library(gt)
  library(arrow)
  library(fs)
  library(patchwork)
})
```

```{r}
#| label: Directories

dir_transformed <- path("data", "transformed")
dir_derived <- path("data", "derived")

determinants_path <- path(
  dir_derived,
  "climate_determinants.parquet"
)
```

```{r}
#| label: Control

# Recalculate the determinant dataset from the transformed ERA5 climatologies.
#
# Set to TRUE after the transformed climate data have changed or when the
# determinant dataset needs to be rebuilt. After a successful calculation,
# FALSE can be used for fast previewing and rendering.
run_determinant_calculation <- FALSE
```

## Climate Periods

The analysis uses three non-overlapping climate intervals to provide a consistent framework for comparing long-term and recent climate conditions and their resulting classifications.

```{r}
#| label: climate-periods

source("R/climate-periods.R")

climate_periods |>
  transmute(
    `Period` = period_name,
    `Short Label` = period_label,
    `Abbreviation` = period_abbr,
    `Years` = sprintf("%d–%d", start_year, end_year)
  ) |>
  gt()

period_labels <- setNames(
  paste0(
    climate_periods$period_label,
    " (",
    climate_periods$start_year,
    "–",
    climate_periods$end_year,
    ")"
  ),
  climate_periods$period_abbr
)

period_palette <- setNames(
  viridisLite::viridis(
    nrow(climate_periods),
    option = "D",
    begin = 0.25,
    end = 0.80
  ),
  climate_periods$period_abbr
)
```

## Determinants

The determinant calculations are applied independently to each climate interval.

### Temperature

- **TAVG** — annual average temperature (°C)
- **TMAX** — monthly maximum temperature (°C)
- **TMIN** — monthly minimum temperature (°C)
- **TG10** — number of months with average temperature ≥ 10 °C

### Precipitation

- **PTOT** — annual total precipitation (mm)
- **PWNT** — winter total precipitation (mm)
- **PSMR** — summer total precipitation (mm)
- **PMIN** — monthly minimum precipitation (mm)
- **PL60** — number of months with total precipitation < 60 mm

### Aridity

- **ATHR** — aridity threshold (cm)

The aridity threshold is calculated as:

**R = 2.3T − 0.64Pw + 41**

where **R** is the aridity threshold in centimetres, **T** is average annual air temperature in °C, and **Pw** is the percentage of annual precipitation occurring in winter.

Reference: <https://www.int-res.com/articles/cr_oa/c059p001.pdf>

```{r}
#| label: determinant-definitions

t_determinants <- c(
  "TAVG",
  "TMAX",
  "TMIN",
  "TG10"
)

p_determinants <- c(
  "PTOT",
  "PWNT",
  "PSMR",
  "PMIN",
  "PL60"
)

a_determinants <- "ATHR"

aridity_variables <- c(
  "wpct",
  "trm1",
  "trm2",
  "trm3"
)

determinants <- c(
  t_determinants,
  p_determinants,
  aridity_variables,
  a_determinants
)

determinant_metadata <- tribble(
  ~Determinant, ~Description,                                      ~Role,                   ~Group,
  "TAVG",       "Average Annual Temperature (°C)",                 "Primary Determinant",  "Temperature",
  "TMAX",       "Monthly Maximum Temperature (°C)",                "Primary Determinant",  "Temperature",
  "TMIN",       "Monthly Minimum Temperature (°C)",                "Primary Determinant",  "Temperature",
  "TG10",       "Months with Average Temperature ≥ 10°C",          "Derived Determinant",  "Temperature",
  "PTOT",       "Annual Total Precipitation (mm)",                 "Primary Determinant",  "Precipitation",
  "PWNT",       "Winter Total Precipitation (mm)",                 "Primary Determinant",  "Precipitation",
  "PSMR",       "Summer Total Precipitation (mm)",                 "Primary Determinant",  "Precipitation",
  "PMIN",       "Monthly Minimum Precipitation (mm)",              "Primary Determinant",  "Precipitation",
  "PL60",       "Months with Total Precipitation < 60 mm",         "Derived Determinant",  "Precipitation",
  "wpct",       "Winter % of Total Precipitation (%)",             "Intermediate Metric",  "Aridity",
  "trm1",       "Term 1: TAVG × 2.3",                              "Formula Term",          "Aridity",
  "trm2",       "Term 2: Winter Precipitation % × 0.64",           "Formula Term",          "Aridity",
  "trm3",       "Term 3: Constant 41",                             "Formula Term",          "Aridity",
  "ATHR",       "Aridity Threshold (cm): Term 1 − Term 2 + Term 3","Derived Determinant",  "Aridity"
)
```

## Functions

```{r}
#| label: functions

fn_t_determinants <- function(df) {
  month_cols <- paste0("m", 1:12)

  df |>
    mutate(
      TAVG = rowMeans(
        across(all_of(month_cols)),
        na.rm = TRUE
      ),
      TMAX = do.call(
        pmax,
        c(
          across(all_of(month_cols)),
          na.rm = TRUE
        )
      ),
      TMIN = do.call(
        pmin,
        c(
          across(all_of(month_cols)),
          na.rm = TRUE
        )
      ),
      TG10 = rowSums(
        across(
          all_of(month_cols),
          ~ .x >= 10
        ),
        na.rm = TRUE
      ),
      TG10 = as.integer(TG10)
    ) |>
    select(-all_of(month_cols))
}

fn_p_determinants <- function(df) {
  month_cols <- paste0("m", 1:12)

  df |>
    mutate(
      PTOT = rowSums(
        across(all_of(month_cols)),
        na.rm = TRUE
      ),
      # Winter: October–March in the Northern Hemisphere and
      # April–September in the Southern Hemisphere.
      PWNT = case_when(
        lat >= 0 ~ m1 + m2 + m3 + m10 + m11 + m12,
        lat < 0 ~ m4 + m5 + m6 + m7 + m8 + m9
      ),
      PSMR = case_when(
        lat >= 0 ~ m4 + m5 + m6 + m7 + m8 + m9,
        lat < 0 ~ m1 + m2 + m3 + m10 + m11 + m12
      ),
      PMIN = do.call(
        pmin,
        c(
          across(all_of(month_cols)),
          na.rm = TRUE
        )
      ),
      PL60 = rowSums(
        across(
          all_of(month_cols),
          ~ .x < 60
        ),
        na.rm = TRUE
      ),
      PL60 = as.integer(PL60)
    ) |>
    select(-all_of(month_cols))
}

fn_aridity_threshold <- function(df) {
  df |>
    mutate(
      trm1 = 2.3 * TAVG,
      wpct = if_else(
        PTOT != 0,
        (PWNT / PTOT) * 100,
        0
      ),
      trm2 = 0.64 * wpct,
      trm3 = 41,
      ATHR = trm1 - trm2 + trm3
    )
}

fn_process_period <- function(period_row) {
  start_year <- period_row$start_year[[1]]
  end_year <- period_row$end_year[[1]]

  temperature_path <- path(
    dir_transformed,
    sprintf(
      "t_agg_%d-%d.parquet",
      start_year,
      end_year
    )
  )

  precipitation_path <- path(
    dir_transformed,
    sprintf(
      "p_agg_%d-%d.parquet",
      start_year,
      end_year
    )
  )

  required_files <- c(
    temperature_path,
    precipitation_path
  )

  if (!all(file_exists(required_files))) {
    stop(
      "Missing transformed climate data for ",
      period_row$period_abbr[[1]],
      ". Run 02-data-transformation.qmd first."
    )
  }

  df_t_det <- read_parquet(
    temperature_path
  ) |>
    fn_t_determinants()

  df_p_det <- read_parquet(
    precipitation_path
  ) |>
    fn_p_determinants()

  df_t_det |>
    left_join(
      df_p_det,
      by = c(
        "lon",
        "lat",
        "area",
        "lsm"
      )
    ) |>
    fn_aridity_threshold() |>
    mutate(
      period_id = period_row$period_id[[1]],
      period_name = period_row$period_name[[1]],
      period_label = period_row$period_label[[1]],
      period_abbr = period_row$period_abbr[[1]],
      start_year = start_year,
      end_year = end_year,
      .before = 1
    )
}

fn_weighted_average <- function(df, determinant_names) {
  df |>
    pivot_longer(
      cols = all_of(determinant_names),
      names_to = "Determinant",
      values_to = "value"
    ) |>
    filter(
      is.finite(value),
      is.finite(area),
      area > 0
    ) |>
    group_by(
      period_id,
      period_name,
      period_label,
      period_abbr,
      Determinant
    ) |>
    summarise(
      wgtavg = weighted.mean(
        value,
        w = area
      ),
      .groups = "drop"
    ) |>
    mutate(
      Determinant = factor(
        Determinant,
        levels = determinant_names
      )
    ) |>
    arrange(
      Determinant,
      factor(
        period_abbr,
        levels = climate_periods$period_abbr
      )
    )
}

fn_summary_table <- function(
  weighted_summary,
  determinant_names,
  title,
  include_determinant = TRUE
) {
  table_data <- weighted_summary |>
    filter(
      Determinant %in% determinant_names
    ) |>
    mutate(
      Determinant = as.character(Determinant)
    ) |>
    left_join(
      determinant_metadata,
      by = "Determinant"
    ) |>
    select(
      Determinant,
      Description,
      Role,
      period_abbr,
      wgtavg
    ) |>
    pivot_wider(
      names_from = period_abbr,
      values_from = wgtavg
    ) |>
    select(
      Description,
      all_of(climate_periods$period_abbr),
      Role,
      Determinant
    )

  if (!include_determinant) {
    table_data <- table_data |>
      select(-Determinant)
  }

  table_data |>
    gt() |>
    tab_header(
      title = title,
      subtitle = "Area-Weighted Averages"
    ) |>
    fmt_number(
      columns = all_of(climate_periods$period_abbr),
      decimals = 2
    )
}

fn_build_determinants_plot <- function(df, type) {
  if (type == "t") {
    determinant_names <- c(
      "TAVG",
      "TMAX",
      "TMIN"
    )
    plot_title <- "Temperature Determinants"
    plot_subtitle <- "Annual Average, Monthly Maximum, Monthly Minimum"
    units <- "°C"
  } else if (type == "p") {
    determinant_names <- c(
      "PTOT",
      "PWNT",
      "PSMR"
    )
    plot_title <- "Precipitation Determinants"
    plot_subtitle <- "Annual and Seasonal Totals"
    units <- "mm"
  } else if (type == "a") {
    determinant_names <- c(
      "trm1",
      "trm2",
      "trm3"
    )
    plot_title <- "Aridity Threshold Calculation"
    plot_subtitle <- "Formula Terms"
    units <- "cm"
  } else {
    stop("Unknown determinant plot type: ", type)
  }

  plot_data <- df |>
    filter(
      Determinant %in% determinant_names
    ) |>
    mutate(
      Determinant = factor(
        as.character(Determinant),
        levels = determinant_names
      ),
      period_abbr = factor(
        period_abbr,
        levels = climate_periods$period_abbr
      )
    )

  p <- ggplot(
    plot_data,
    aes(
      x = Determinant,
      y = wgtavg,
      fill = period_abbr
    )
  ) +
    geom_col(
      position = position_dodge(width = 0.75),
      width = 0.65
    ) +
    scale_fill_manual(
      values = period_palette,
      breaks = climate_periods$period_abbr,
      labels = period_plot_labels,
      guide = guide_legend(
        nrow = 1,
        byrow = TRUE
      )
    ) +
    labs(
      title = plot_title,
      subtitle = plot_subtitle,
      x = NULL,
      y = units,
      fill = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "bottom",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(size = 11)
    )

  if (type == "a") {
    p <- p +
      scale_x_discrete(
        labels = c(
          trm1 = "2.3T",
          trm2 = "0.64Pw",
          trm3 = "C"
        )
      )
  }

  p
}

fn_build_single_plot <- function(df, determinant_name) {
  if (determinant_name == "PMIN") {
    plot_subtitle <- "Monthly Minimum"
    units <- "mm"
  } else if (determinant_name == "ATHR") {
    plot_subtitle <- "2.3T − 0.64Pw + C"
    units <- "cm"
  } else {
    stop("Unknown single determinant: ", determinant_name)
  }

  plot_data <- df |>
    filter(
      Determinant == determinant_name
    ) |>
    mutate(
      period_abbr = factor(
        period_abbr,
        levels = climate_periods$period_abbr
      )
    )

  ggplot(
    plot_data,
    aes(
      x = Determinant,
      y = wgtavg,
      fill = period_abbr
    )
  ) +
    geom_col(
      position = position_dodge(width = 0.75),
      width = 0.65
    ) +
    scale_fill_manual(
      values = period_palette,
      breaks = climate_periods$period_abbr,
      labels = period_plot_labels,
      guide = guide_legend(
        nrow = 1,
        byrow = TRUE
      )
    ) +
    labs(
      subtitle = plot_subtitle,
      x = NULL,
      y = units,
      fill = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "bottom",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(size = 11)
    )
}

fn_build_counts_plot <- function(
  df,
  determinant_name,
  plot_subtitle
) {
  plot_data <- df |>
    filter(
      lsm > 0
    ) |>
    transmute(
      period_abbr,
      area,
      value = .data[[determinant_name]]
    ) |>
    group_by(
      period_abbr,
      value
    ) |>
    summarise(
      total_area = sum(
        area,
        na.rm = TRUE
      ),
      .groups = "drop"
    ) |>
    group_by(
      period_abbr
    ) |>
    mutate(
      prop_area = total_area /
        sum(total_area)
    ) |>
    ungroup() |>
    mutate(
      period_abbr = factor(
        period_abbr,
        levels = climate_periods$period_abbr
      )
    )

  ggplot(
    plot_data,
    aes(
      x = value,
      y = prop_area,
      fill = period_abbr
    )
  ) +
    geom_col(
      position = "dodge",
      width = 0.7
    ) +
    scale_y_continuous(
      labels = scales::percent_format(
        accuracy = 1
      )
    ) +
    scale_fill_manual(
      values = period_palette,
      breaks = climate_periods$period_abbr,
      labels = period_plot_labels,
      guide = guide_legend(
        nrow = 1,
        byrow = TRUE
      )
    ) +
    labs(
      subtitle = plot_subtitle,
      x = determinant_name,
      y = "Land Area %",
      fill = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      legend.position = "bottom",
      panel.grid.major.x = element_blank(),
      axis.title.x = element_text(size = 11.5)
    )
}
```

## Main Process

```{r}
#| label: main-process
#| results: hide

if (run_determinant_calculation) {
  determinant_results <- bind_rows(
    lapply(
      seq_len(nrow(climate_periods)),
      function(i) {
        fn_process_period(
          climate_periods[i, ]
        )
      }
    )
  )

  write_parquet(
    determinant_results,
    determinants_path
  )
} else {
  if (!file_exists(determinants_path)) {
    stop(
      "The determinant dataset does not exist. ",
      "Set run_determinant_calculation <- TRUE and run the notebook once."
    )
  }

  determinant_results <- read_parquet(
    determinants_path
  )
}

weighted_summary <- fn_weighted_average(
  determinant_results,
  determinants
)
```

## Determinant Summary Tables

The tables compare area-weighted determinant values across the REF, CSN, and RCI climate intervals.

```{r}
#| label: temperature-summary-table
#| echo: false

fn_summary_table(
  weighted_summary,
  t_determinants,
  "Temperature Determinants"
)
```

```{r}
#| label: precipitation-summary-table
#| echo: false

fn_summary_table(
  weighted_summary,
  p_determinants,
  "Precipitation Determinants"
)
```

```{r}
#| label: aridity-summary-table
#| echo: false

aridity_summary_names <- c(
  "TAVG",
  "PWNT",
  "PTOT",
  "wpct",
  "trm1",
  "trm2",
  "trm3",
  "ATHR"
)

fn_summary_table(
  weighted_summary,
  aridity_summary_names,
  "Aridity Threshold Calculation Detail"
)
```

## Visualizations

The visualizations compare the area-weighted determinants and land-area distributions among the three climate intervals.


```{r}
#| label: plot-labels
period_plot_labels <- setNames(
  paste0(
    climate_periods$period_abbr,
    " (",
    climate_periods$start_year,
    "–",
    climate_periods$end_year,
    ")"
  ),
  climate_periods$period_abbr
)
```

### Temperature

```{r}
#| label: temperature-visualization
#| fig-height: 7
#| out-width: 100%

temperature_determinants_plot <- fn_build_determinants_plot(
  weighted_summary,
  "t"
)

temperature_counts_plot <- fn_build_counts_plot(
  determinant_results,
  "TG10",
  "Months per Year ≥ 10°C"
)

temperature_determinants_plot /
  temperature_counts_plot +
  plot_layout(
    guides = "collect"
  )  &
  theme(
    legend.position = "bottom"
  )
```

### Precipitation

```{r}
#| label: precipitation-visualization
#| fig-height: 8
#| out-width: 100%

precipitation_determinants_plot <- fn_build_determinants_plot(
  weighted_summary,
  "p"
)

precipitation_minimum_plot <- fn_build_single_plot(
  weighted_summary,
  "PMIN"
)

precipitation_counts_plot <- fn_build_counts_plot(
  determinant_results,
  "PL60",
  "Months per Year < 60 mm"
)

precipitation_determinants_plot +
  precipitation_minimum_plot +
  precipitation_counts_plot +
  plot_layout(
    design = "AAB\nCCC",
    guides = "collect"
  ) &
  theme(
    legend.position = "bottom"
  )
```

### Aridity Threshold

```{r}
#| label: aridity-visualization
#| fig-height: 4
#| out-width: 100%

aridity_terms_plot <- fn_build_determinants_plot(
  weighted_summary,
  "a"
)

aridity_threshold_plot <- fn_build_single_plot(
  weighted_summary,
  "ATHR"
)

aridity_terms_plot +
  aridity_threshold_plot +
  plot_layout(
    widths = c(3, 2),
    guides = "collect"
  ) &
  theme(
    legend.position = "bottom"
  )
```
