---
title: "Transition Mapping"
author: "Rick Forest"
date: today

execute:
  echo: true
  warning: false
  message: false

format:
  html:
    toc: true
    number-sections: false
    code-fold: true
    code-summary: "Show code"
    code-tools: true
---

**Objective:** Map the geographic distribution of Köppen–Trewartha climate-classification transitions among the REF, CSN, and RCI climate intervals. This notebook shows where climate types changed, distinguishes within-group type changes from broader climate-group changes, examines the expansion of dry and desert climate classifications and the retreat of polar classifications, and provides a long-term spatial summary of change from REF to RCI.

The transition-analysis notebook quantified how much land changed classification and identified the most important transitions. This final stage adds the geographic dimension: **where did those changes occur?**

## Initialization

```{r}
#| label: Libraries

suppressPackageStartupMessages({
  library(tidyverse)
  library(sf)
  library(gt)
  library(arrow)
  library(fs)
  library(patchwork)
})
```

```{r}
#| label: Directories

dir_reference <- path("data", "reference")
dir_derived <- path("data", "derived")

transitions_path <- path(
  dir_derived,
  "climate_transitions.parquet"
)

group_codes_path <- path(
  dir_reference,
  "ktc_group_codes.csv"
)

type_codes_path <- path(
  dir_reference,
  "ktc_type_codes.csv"
)

coastline_path <- path(
  dir_reference,
  "global-coastline.gpkg"
)
```

## Climate Periods

```{r}
#| label: climate-periods

source("R/climate-periods.R")

climate_periods |>
  transmute(
    `Period` = period_name,
    `Short Label` = period_label,
    `Abbreviation` = period_abbr,
    `Years` = sprintf("%d–%d", start_year, end_year)
  ) |>
  gt()

transition_comparisons <- tribble(
  ~comparison_id, ~comparison_label, ~from_period_id,     ~to_period_id,
  "ref_csn",      "REF → CSN",       "reference_period", "standard_normal",
  "csn_rci",      "CSN → RCI",       "standard_normal",  "recent_interval",
  "ref_rci",      "REF → RCI",       "reference_period", "recent_interval"
)

comparison_levels <- transition_comparisons$comparison_label
```

## Mapping Reference

```{r}
#| label: mapping-reference

required_reference_files <- c(
  group_codes_path,
  type_codes_path,
  coastline_path
)

if (!all(file_exists(required_reference_files))) {
  stop(
    "Missing transition-mapping reference files in data/reference. ",
    "Run 01-data-acquisition.qmd first."
  )
}

group_codes <- read_csv(
  group_codes_path,
  show_col_types = FALSE
)

type_codes <- read_csv(
  type_codes_path,
  show_col_types = FALSE
)

group_levels <- group_codes$ktc
type_levels <- type_codes$ktc

group_palette <- setNames(
  group_codes$hex,
  group_codes$ktc
)

type_palette <- setNames(
  type_codes$hex,
  type_codes$ktc
)

sf_coastline <- st_read(
  coastline_path,
  quiet = TRUE
)

rbn_prj <- "ESRI:54030"

change_palette <- c(
  "Type change within group" = "#8C9AA6",
  "Climate group change" = "#9B5E6E"
)
```

## Functions

```{r}
#| label: functions

fn_transition_points <- function(df) {
  st_as_sf(
    df,
    coords = c(
      "lon",
      "lat"
    ),
    crs = 4326,
    remove = FALSE
  )
}

fn_comparison_subtitle <- function(comparison_id) {
  comparison_row <- transition_comparisons |>
    filter(
      .data$comparison_id ==
        .env$comparison_id
    )

  if (nrow(comparison_row) != 1) {
    stop(
      "Expected exactly one transition comparison for ",
      comparison_id,
      "."
    )
  }

  comparison_row$comparison_label[[1]]
}

fn_build_overall_change_map <- function(
  transitions_df,
  comparison_id
) {
  plot_df <- transitions_df |>
    filter(
      .data$comparison_id ==
        .env$comparison_id,
      type_changed
    ) |>
    mutate(
      change_level = if_else(
        group_changed,
        "Climate group change",
        "Type change within group"
      ),
      change_level = factor(
        change_level,
        levels = c(
          "Type change within group",
          "Climate group change"
        )
      )
    ) |>
    fn_transition_points()

  ggplot() +
    geom_sf(
      data = plot_df,
      aes(
        color = change_level
      ),
      size = 0.08,
      alpha = 0.85
    ) +
    geom_sf(
      data = sf_coastline,
      color = "grey35",
      linewidth = 0.15
    ) +
    scale_color_manual(
      values = change_palette,
      drop = FALSE
    ) +
    guides(
      color = guide_legend(
        override.aes = list(
          size = 5,
          alpha = 1
        )
      )
    ) +
    coord_sf(
      crs = rbn_prj,
      datum = NA
    ) +
    labs(
      title = "Geographic Distribution of KTC Classification Change",
      subtitle = fn_comparison_subtitle(
        comparison_id
      ),
      color = NULL
    ) +
    theme_void() +
    theme(
      plot.title = element_text(
        face = "bold"
      ),
      legend.position = "bottom",
      legend.text = element_text(
        size = 12
      )
    )
}

fn_build_themed_transition_map <- function(
  transitions_df,
  comparison_id,
  filter_expression,
  color_variable,
  color_levels,
  palette,
  title,
  subtitle,
  legend_title
) {
  plot_df <- transitions_df |>
    filter(
      .data$comparison_id ==
        .env$comparison_id
    ) |>
    filter(
      {{ filter_expression }}
    ) |>
    mutate(
      map_class = factor(
        .data[[color_variable]],
        levels = color_levels
      )
    ) |>
    fn_transition_points()

  ggplot() +
    geom_sf(
      data = plot_df,
      aes(
        color = map_class
      ),
      size = 0.12,
      alpha = 0.9
    ) +
    geom_sf(
      data = sf_coastline,
      color = "grey35",
      linewidth = 0.15
    ) +
    scale_color_manual(
      values = palette,
      drop = TRUE
    ) +
    guides(
      color = guide_legend(
        override.aes = list(
          size = 5,
          alpha = 1
        )
      )
    ) +
    coord_sf(
      crs = rbn_prj,
      datum = NA
    ) +
    labs(
      title = title,
      subtitle = paste(
        fn_comparison_subtitle(
          comparison_id
        ),
        "—",
        subtitle
      ),
      color = legend_title
    ) +
    theme_void() +
    theme(
      plot.title = element_text(
        face = "bold"
      ),
      legend.position = "bottom",
      legend.title = element_text(
        size = 12
      ),
      legend.text = element_text(
        size = 11
      )
    )
}

fn_build_long_term_group_map <- function(
  transitions_df
) {
  plot_df <- transitions_df |>
    filter(
      comparison_id == "ref_rci",
      group_changed
    ) |>
    mutate(
      destination_group = factor(
        to_grp,
        levels = group_levels
      )
    ) |>
    fn_transition_points()

  ggplot() +
    geom_sf(
      data = plot_df,
      aes(
        color = destination_group
      ),
      size = 0.09,
      alpha = 0.9
    ) +
    geom_sf(
      data = sf_coastline,
      color = "grey35",
      linewidth = 0.15
    ) +
    scale_color_manual(
      values = group_palette,
      breaks = group_levels,
      drop = FALSE
    ) +
    guides(
      color = guide_legend(
        override.aes = list(
          size = 5,
          alpha = 1
        )
      )
    ) +
    coord_sf(
      crs = rbn_prj,
      datum = NA
    ) +
    labs(
      title = "Long-Term KTC Climate Group Transitions",
      subtitle = "REF → RCI — changed grid cells colored by destination climate group",
      color = "Destination Group"
    ) +
    theme_void() +
    theme(
      plot.title = element_text(
        face = "bold"
      ),
      legend.position = "bottom",
      legend.title = element_text(
        size = 12
      ),
      legend.text = element_text(
        size = 11
      )
    )
}
```

## Import

```{r}
#| label: import

if (!file_exists(transitions_path)) {
  stop(
    "The transition dataset does not exist. ",
    "Run 05-transition-analysis.qmd first."
  )
}

transition_results <- read_parquet(
  transitions_path
)
```

## Overall Spatial Pattern of Change

These maps show only grid cells whose KTC **type** changed. Cells that remained within the same broad climate group are distinguished from cells that crossed a climate-group boundary. Unchanged land is omitted so that the spatial pattern of change is not overwhelmed by the much larger stable classification area.

```{r}
#| label: overall-change-maps
#| fig-width: 16
#| fig-height: 18
#| out-width: 100%
#| column: screen-inset

overall_ref_csn <- fn_build_overall_change_map(
  transition_results,
  "ref_csn"
)

overall_csn_rci <- fn_build_overall_change_map(
  transition_results,
  "csn_rci"
)

overall_ref_rci <- fn_build_overall_change_map(
  transition_results,
  "ref_rci"
)

overall_ref_csn /
  overall_csn_rci /
  overall_ref_rci +
  plot_layout(
    guides = "collect"
  ) &
  theme(
    legend.position = "bottom"
  )
```

## Expansion of Dry and Desert Climate Classifications

The following maps focus on long-term REF → RCI transitions into increasingly dry Köppen–Trewartha classifications. These maps describe **climate-classification change**, not land degradation or desertification.

### Expansion into Group B — Dry

This map shows grid cells that were not classified as Group B during REF but are classified as Group B during RCI. Changed cells are colored by their **origin climate group**, showing both where dry climates expanded and which broad climate classifications they replaced.

```{r}
#| label: expansion-into-group-b
#| fig-width: 16
#| fig-height: 9
#| out-width: 100%
#| column: screen-inset

fn_build_themed_transition_map(
  transitions_df = transition_results,
  comparison_id = "ref_rci",
  filter_expression =
    from_grp != "B" &
    to_grp == "B",
  color_variable = "from_grp",
  color_levels = group_levels,
  palette = group_palette,
  title = "Expansion into KTC Group B — Dry",
  subtitle = "new Group B grid cells colored by origin climate group",
  legend_title = "Origin Group"
)
```

### Transition into Type BW — Desert

This map isolates the more specific transition into the KTC **BW — Desert (Arid)** climate type. Changed cells are colored by their **origin climate type**, allowing the locations and prior classifications of newly classified desert climates to be examined directly.

```{r}
#| label: transition-into-type-bw
#| fig-width: 16
#| fig-height: 9
#| out-width: 100%
#| column: screen-inset

fn_build_themed_transition_map(
  transitions_df = transition_results,
  comparison_id = "ref_rci",
  filter_expression =
    from_ktc != "BW" &
    to_ktc == "BW",
  color_variable = "from_ktc",
  color_levels = type_levels,
  palette = type_palette,
  title = "Transition into KTC Type BW — Desert",
  subtitle = "new BW grid cells colored by origin climate type",
  legend_title = "Origin Type"
)
```

## Retreat of Polar Climate Classifications

This map shows grid cells classified as **Group F — Polar** during REF that are assigned to a different climate group during RCI. Changed cells are colored by their **destination climate group**, showing where polar classifications retreated and which broad climates replaced them.

```{r}
#| label: retreat-from-group-f
#| fig-width: 16
#| fig-height: 9
#| out-width: 100%
#| column: screen-inset

fn_build_themed_transition_map(
  transitions_df = transition_results,
  comparison_id = "ref_rci",
  filter_expression =
    from_grp == "F" &
    to_grp != "F",
  color_variable = "to_grp",
  color_levels = group_levels,
  palette = group_palette,
  title = "Retreat of KTC Group F — Polar",
  subtitle = "former Group F grid cells colored by destination climate group",
  legend_title = "Destination Group"
)
```

## Long-Term Spatial Summary

The final map emphasizes the broadest project comparison, REF → RCI. It shows only grid cells that changed KTC climate **group**, with each changed cell colored by its destination group. Stable land is omitted so that the spatial structure of long-term classification change remains visible.

```{r}
#| label: long-term-transition-map
#| fig-width: 16
#| fig-height: 9
#| out-width: 100%
#| column: screen-inset

fn_build_long_term_group_map(
  transition_results
)
```

```{r}
#| label: export-github-image
#| echo: false
social_preview_map <- fn_build_long_term_group_map(
  transition_results
)

ggsave(
  "social-preview-map.png",
  plot = social_preview_map,
  width = 16,
  height = 9,
  units = "in",
  dpi = 200,
  bg = "white"
)

```